---
title: "Influence of Sex"
author: "C. Chabbert"
date: "9/5/2018"
output: html_document
---

This script will assess whether Khk expression levels are tissue specific.

```{r setup}
## Libraries
library("DESeq2")
library("tximport")
library("DT")

## Useful functions
Load_QoRTs_counts <- function(decoder, dir, type){
  
  ## Set up sample list together with the directories
  qc.data.dir = as.character(decoder$qc.data.dir)
  names(qc.data.dir) = as.character(decoder$unique.ID)
  
  ## Load counts
  raw.counts.list = lapply(names(qc.data.dir),
                           function(x){ Load_QoRTs_table(x, qc.data.dir, dir, type) })
  
  ## Collapse list into data frame and re-format
  df.counts = dplyr::bind_rows(raw.counts.list) %>%
    tidyr::spread(Sample, Count)
  
  ## Conversion to a DESeq compatible matrix
  mat.counts = as.matrix(df.counts[, 2:ncol(df.counts)])
  colnames(mat.counts) = colnames(df.counts[, 2:ncol(df.counts)])
  rownames(mat.counts) = as.character(df.counts$Gene)
  
  return(mat.counts)

}

Load_QoRTs_table <- function(sample, qc.data.dir, dir, type){
  
  ## Pass the extension name for each file
  f.extension = "_"
  if (type == "DESeq"){
    f.extension = "QC.geneCounts.formatted.for.DESeq.txt.gz"
  }
  else if (type == "DEXSeq"){
    f.extension = "QC.exonCounts.formatted.for.DEXSeq.txt.gz"
  }
  
  ## Create path to load
  path = paste(dir, 
               qc.data.dir[sample], 
               f.extension, 
               sep = "/")
  
  ## Load the data
  df = read.delim(path,
                  header = F, 
                  stringsAsFactors = F,
                  sep = "\t")
  
  ## Addiiton of sample information and column names
  df$Sample = gsub("_qorts", "", qc.data.dir[sample])
  colnames(df) = c("Gene", "Count", "Sample")
  
  return(df)
}
```


```{r data_load}
## Load count tables from QoRTs (they should be DESeq2 compatible)

## Sample annotations
decoder = read.csv("tissue_annotations.csv",
                   stringsAsFactors = FALSE,
                   header = TRUE) %>%
  dplyr::rename(unique.ID = Sample,
                group.ID = Group) %>%
  dplyr::mutate(qc.data.dir = paste0(unique.ID, "_qorts")) %>%
  dplyr::filter(!(group.ID %in% c("testis", "ovary")))

# Directory where reports are stored
directory.qorts = "/qorts/"

# Data load
mat.counts = Load_QoRTs_counts(decoder, directory.qorts, "DESeq")
```


# 1 -- Preparation of the DESeq2 objects

In order to conduct the differential expression analysis, we now prepare the DESeq2 object,
starting with the experimental design and conditions.

```{r design}
## Preparation of design matrix
colData = dplyr::rename(decoder, 
                        Sample = unique.ID, Tissue = group.ID) %>%
  dplyr::select(-qc.data.dir)

rownames(colData) = as.character(colData$Sample)

colData$Sample    = factor(colData$Sample, levels = colData$Sample)
colData$Tissue  = factor(colData$Tissue, 
                         levels = c("adrenal_gland", "bone_marrow", "brain", "forestomach",
                                    "heart", "kidney", "large_intestine", "liver",
                                    "lung", "muscle", "small_intestine",
                                    "spleen", "stomach", "thymus"))
colData$Sex = factor(colData$Sex,
                    levels = c("female", "male"))

# We create an annotation df for plotting functions
annotations.df = colData
```

As QoRTs includes some "non-gene" counts (for ex. multiple alignments), we need
to filter these out before proceding with DESeq.

```{r mat_cleanup}
## row ids
ids = grep("ENS", rownames(mat.counts), invert = TRUE)

## Display results
print(mat.counts[ids,])

## Matrix cleanup
mat.counts.all = mat.counts
mat.counts = mat.counts[, rownames(colData)]
mat.counts = mat.counts[-ids, row.names(colData)]
```

The next step will be to clean up the gene list to remove rRNA, tRNAs,
pseudo genes and other transcript categories which cannot be captured by
the protocol.

```{r gene_cleanup}
toExclude = c(ig_variable_chains, ig_pseudogenes, vault,
              nmd_degradation, non_coding_other, pseudogenes, lincRNA, others)
gtf = rtracklayer::import(con = "/mouse/M14_GRCm38.p5/gencode.vM14.primary_assembly.annotation.gtf.gz")
genes.to.keep = data.frame(gtf[, c("type", "gene_id", "gene_name", "gene_type")]) %>%
  dplyr::filter(type == "gene") %>%
  dplyr::filter(!(gene_type %in% toExclude))

mat.counts = mat.counts[as.character(genes.to.keep$gene_id),]
```

We can now create a DESeq object. In order to simplify the first analysis,
we create a "Group" variable which will describe both the genotype and the condition.
The introduction of a proper formula with interaction terms may be checked later on
when testing for the effect on condition or of genotype using a Likelihood Ratio.

```{r deseq_object}
## Creation of the DESeq2 objects
dds = DESeqDataSetFromMatrix(countData = mat.counts,
                             colData = colData,
                             design = ~ Tissue + Tissue:Sex)
```

# 2 -- Differential gene expression analysis

Before running the dge analysis, we first remove genes with very low couns
across all samples (threshold set at 50)

```{r size_factors}
## Check how many genes have less than 20 supporting reads
sum(rowSums(counts(dds)) < 50)

## These genes are removed from downstreem analysis
dds <- dds[rowSums(counts(dds)) > 50,]

## We now estimate the size factors
dds <- estimateSizeFactors(dds)
```

We examine the size factors:

```{r sizeFactors_plot, fig.align = "center", fig.width = 10, fig.height = 6}
annotations.df$SizeFactors = sizeFactors(dds)[row.names(annotations.df)]

ggplot(annotations.df, aes(Sample, SizeFactors)) +
  geom_point(aes(col = Tissue), size = 5) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

We use a LRT approach to evaluate which genes have a sex specific expression
in each tissue. The output is saved as an RData object.

```{r}
dds.LRT <- DESeq(dds, test="LRT", reduced=~Tissue)
plotDispEsts(dds.LRT)
res.LRT <- results(dds.LRT)
save(dds.LRT,
     file = "dds.LRT.li.tissues.no_testis.wSex.RData")
```

We display the results for the Khk genes:

```{r}
res.LRT["ENSMUSG00000029162.15", ]
```


```{r}
sessionInfo()
```
